# An introduction to plotting using ggplot2

```{r, echo=F}
data <- read.csv("./Data/China_dataset.csv")
```

The package `ggplot2` is widely used for data visualization in R. This package is extremely powerful. I used to like using basic R code for plotting but eventually, I had to admit that ggplot is extremely cool and had to adopt it. 

`ggplot2`is based on the grammar of graphics, which allows users to create complex plots from data in a systematic way.

As with any R package, before you can use ggplot2, you need to install it (if you haven't already) and load it into your R session.

```{r}
#install.packages("ggplot2")
library(ggplot2)
```

## Basic Concepts

There are a few concepts that we need to know to understand how to code ggplots. 

Data: The dataset you want to visualize.
Aesthetics (aes): The visual properties of the plot (e.g., x and y position, color, size).
Geometries (geom_): The actual marks we put on the plot (e.g., points, lines, bars). We can, for instance, use geom_line() for plotting lines.
Facets: Subplots that display subsets of the data.
Scales: Control how data values are mapped to visual properties.
Themes: Control the overall appearance of the plot (e.g., font size, background color).

## Creating Your First Plot

Let’s start with a simple scatter plot using the dataset we have been working with. 
### A basic scatter plot

First, lets filter the data to keep only year 2015. Let's review some data manipulation steps:

```{r}
# Create data column
data$Date <- as.Date(data$Date, format = "%d-%M-%Y")
# Create year column and call the new object data2
data2<- data %>%
	mutate(Year = format(Date, "%Y")) 
# Filter year 2015
data3 <- data2 |> filter(Year == 2015)
```

Now, we can create our first plot. Lets plot temperature against dissolved oxigen

```{r}
ggplot(data = data3, aes(x = Temperature..cel., y = Dissolved.Oxygen..mg.l.)) +
  geom_point()
```

In here, `ggplot(data = data3, aes(x = Temperature..cel., y = Dissolved.Oxygen..mg.l.)) ` initializes the plot with the `data3` dataset, setting `Temperature..cel.` on the x-axis and `Dissolved.Oxygen..mg.l.`. `geom_point()` adds points to the plot to create a scatter plot.

We can see that as temperature increases, the amount of oxygen dissolved in the water decreases.

We can also see a line of dots that give you a hint that something may be wrong with the measures. But for the purpose of this practical, we can ignore that.

### Customizing the plot

#### Titles and labels
We now have a basic plot. Let's start to customize it. We will first add a title, subtitle, and axis labels with the `labs()` function.

```{r}
ggplot(data = data3, aes(x = Temperature..cel., y = Dissolved.Oxygen..mg.l.)) +
  geom_point() +
  labs(title = "Scatter Plot of temp vs diss. oxygen",
       x = "Temperature (C)",
       y = "Dissolved oxygen (mg/l)")
```

#### Color and size

We can also change the point color using `col = "red"`. Try other colors, like "blue", or "green".

```{r}
ggplot(data = data3, aes(x = Temperature..cel., y = Dissolved.Oxygen..mg.l.)) +
  geom_point(col = "red") +
  labs(title = "Scatter Plot of temp vs diss. oxygen",
       x = "Temperature (C)",
       y = "Dissolved oxygen (mg/l)")
```

Looking pretty good. 

#### Linear trends

What if we want to add a line trend?

We can use the `geom_smooth()` function to add a linear trend. Method refers to the moothing method (function). Here we use a linear model. se controls whether you plot the standard error. And color controls the color of the line.

```{r}
ggplot(data = data3, aes(x = Temperature..cel., y = Dissolved.Oxygen..mg.l.)) +
  geom_point(col = "red") +
  labs(title = "Scatter Plot of temp vs diss. oxygen",
       x = "Temperature (C)",
       y = "Dissolved oxygen (mg/l)") +
  geom_smooth(method = "lm", se = T, color = "blue")
```

We now can clearly see the negative relationship.

### Using Different Geometries

Let's try now different geometries.

#### Histogram

For a histogram we use `geom_histogram()`.

```{r}
ggplot(data = data3, aes(x = Temperature..cel.)) +
  geom_histogram(binwidth = 2, fill = "blue", color = "black") +
  labs(title = "Histogram of water temperature",
       x = "Temperature (C)",
       y = "Count")
```

#### Box Plot

We use `geom_box()` for a boxplot.

This can be useful to visualize the distribution of a variable across a different categories. For instance, we can look at the concentration of nitrogen across different water quality indexes:

```{r}
ggplot(data = data3, aes(x = factor(CCME_WQI), y = Nitrogen..mg.l.)) +
  geom_boxplot() +
  labs(x = "Water Quality Index",
       y = "Nitrogen (mg/l)")
```

### Customizing the plot appearance

#### Themes

The `themes` control the overall look of your plot. There are many themes available. For isntance, we can use `theme_minimal()`.

```{r}
ggplot(data = data3, aes(x = factor(CCME_WQI), y = Nitrogen..mg.l.)) +
  geom_boxplot() +
  labs(x = "Water Quality Index",
       y = "Nitrogen (mg/l)") +
  theme_minimal()
```

Other available themes include `theme_gray()`, `theme_classic()`, `theme_bw()`, and more. Check them out yourself.

With ggplot2, you have a powerful tool to explore and present your data in compelling ways.

There are many options that you can control on ggplot. The best way to learn all the possibilities is by playing with it. Pretty much, everything can be customized. 
Feel free to experiment with different datasets and ggplot2 functions to create the visualizations that best communicate your insights!

### Saving Your Plot

To save your plot to a file use the function `ggsave()`. Note that first you need to save your plot as an object.

```{r, eval=FALSE}
p <- ggplot(data = mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  labs(title = "Scatter Plot of MPG vs Weight")

ggsave("scatter_plot.png", plot = p, width = 6, height = 4)
```

### A time series plot

Finally, we want to explore how different parameters change over time.

For this, we need to combine everything we learned to first create the variables we need and then plot them. A full analysis.

First, lets look at the `data2` dataframe we created by using the `summary()` funtion:

```{r}
summary(data2)
```
Notice that there are many dates missing in the dataset. That will create problems, so we need to get rid of those rows. For that, we can use the function `filter()` and keep just data without missing dates using `complete.cases(Date)`:

```{r}
nrow(data2)
data2 <- data2 |> filter(complete.cases(Date))
nrow(data2)
```

Note, we have removed the 500 rows with NAs.

Now, we can create a new dataframe with the average value per year across all years:

```{r}
China <- data2 |>
	mutate(Year = format(Date, "%Y")) |>
  group_by(Year) %>%
  summarise(
    Ammonia = mean(Ammonia..mg.l., na.rm = TRUE),             
    Dissolved.Oxygen = mean(Dissolved.Oxygen..mg.l., na.rm = TRUE),         
    pH = mean(pH..ph.units., na.rm = TRUE),
    Temperature = mean(Temperature..cel., na.rm = TRUE),             
    Nitrogen = mean(Nitrogen..mg.l., na.rm = TRUE),
    Nitrate = mean(Nitrate..mg.l., na.rm = TRUE),
    .groups = "drop"
  )

China$Year <- as.Date(strptime(China$Year, "%Y")) # Convert year back to date format
```

Finally, we can create time series to see how these parameters have changed across years:

```{r}
ggplot(China, aes(x=Year, y = Ammonia)) + geom_point() + geom_line(aes(group = "1")) + geom_smooth(method = "gam", se = T, color = "blue")
ggplot(China, aes(x=Year, y = Dissolved.Oxygen)) + geom_point() + geom_line(aes(group = "1")) + geom_smooth(method = "gam", se = T, color = "blue")
ggplot(China, aes(x=Year, y = pH)) + geom_point() + geom_line(aes(group = "1")) + geom_smooth(method = "gam", se = T, color = "blue")
ggplot(China, aes(x=Year, y = Temperature)) + geom_point() + geom_line(aes(group = "1")) + geom_smooth(method = "gam", se = T, color = "blue")
ggplot(China, aes(x=Year, y = Nitrogen)) + geom_point() + geom_line(aes(group = "1")) + geom_smooth(method = "gam", se = T, color = "blue")
ggplot(China, aes(x=Year, y = Nitrate)) + geom_point() + geom_line(aes(group = "1")) + geom_smooth(method = "gam", se = T, color = "blue")
```


