# Assignment

This assignment is intended to practice the skills you have learned so far. 

I want you to analyse the trend in different fresh water parameters but this time for a water system in Ireland.

What you need to do is:

1. Start a new R script.

2. Load the dataset: "Ireland_dataset.csv"

```{r}
Ireland <- read.csv("./Data/Ireland_dataset.csv")
```

3. Filter the water Area. Check this table and use the area name based on your group number:

```{r, echo=FALSE}
names <- c(
  "Corrib, Aille",
  "Corrib, Ballyquirke",
  "Corrib, Carra",
  "Corrib, Bofin GY",
  "Corrib, Corrib Lower",
  "Lough Neagh _ Lower Bann, Emy",
  "Foyle, Derg DL",
  "Foyle, Finn DL",
  "Foyle, Mourne DL",
  "Newry, Fane, Glyde and Dee, Brackan",
  "Newry, Fane, Glyde and Dee, Monalty",
  "Newry, Fane, Glyde and Dee, Muckno",
  "Newry, Fane, Glyde and Dee, Naglack",
  "Newry, Fane, Glyde and Dee, Spring",
  "Mal Bay, Doo CE",
  "Mal Bay, Keagh",
  "Mal Bay, Lickeen",
  "Mal Bay, Naminna",
  "Lough Swilly, Akibbon",
  "Lough Swilly, Fern",
  "Lough Swilly, Gartan",
  "Tralee Bay-Feale, Cam KY",
  "Tralee Bay-Feale, Gill KY",
  "Lee, Cork Harbour and Youghal Bay, Allua",
  "Lower Shannon (C), Graney"
)

table_df <- data.frame(
  `Group Number` = 1:25,
  `Area Name` = names
)

knitr::kable(table_df)
```

Hint 1: You need to load the `dpylr` package. 
Hint 2: Make sure the variable Area is in the correct structure, factor. 

4. Similar to the China example, calculate the average value of ammonia, dissolved oxygen, PH, temperature, and nitrogen per year.

Hint: You need to load the `ggplot2` package. 

5. For each variable, create a time series plot showing the trend.

Hint: Adapt the code we used for the time series plot using the data from China.
For more points: Customize your plots, using different colors. Make sure your variables are properly named.

6. In a Word document, paste your entire R code and all your plots. 

Good codes are those easily legible, well organised and well annotated. You will get marks accordingly.

7. Write a short paragraph interpreting the parameter trends in the figures, including a potential explanation for the trends you observe. 

8. Submit on Canvas (Do not forget to write the names of all participants)